\bf Longitudinal Principal Manifold Estimation

### 2024-07-13 paper/main.tex
Updates from Overleaf

\if1\blind
{
  \title{\bf Longitudinal Principal Manifold Estimation}
  \author{
    Robert Zielinski\hspace{.2cm}\\
    Department of Biostatistics, Brown University\\
    Kun Meng \\
    Division of Applied Mathematics, Brown University\\
    and \\
    Ani Eloyan \\
    Department of Biostatistics, Brown University\\
    for the Alzheimer's Disease Neuroimaging Initiative}
  \maketitle
} \fi

@@

Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}. Manifold learning approaches have been applied in various scientific fields, e.g., single-cell biology \citep{ding2023learning} and robotics \citep{gao2023k}. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures. Nonlinear manifold learning methods have been extensively studied in the literature. However, many popular approaches, such as Isomap \citep{tenenbaumGlobalGeometricFramework2000}, locally linear embedding \citep{roweisNonlinearDimensionalityReduction2000}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies. That is, these methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension. %We use the oldest linear manifold learning method, principal component analysis \citep[PCA,][]{Pearson1901on}, as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal PCA method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}. As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space. However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. Additionally, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.

@@

The framework for principal manifolds originated with the concept of principal curves \citep{hastiePrincipalCurves1989}, which are essentially curves that pass through the middle of a data cloud. Motivated by the penalization approach presented by \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework, which extends principal curves to arbitrarily high intrinsic dimensions. In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional spaces, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure 1 in the online supplement. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a $d$-variable function $u$, $\nabla^{\otimes 2} u$ denotes the Hessian matrix of $u$, defined as $r \to \left( \frac{\partial ^2 u}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are defined as follows.
\begin{definition}
  \label{def:principal_manifolds}
  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability distribution $\mathbb{P}$ such that $\mathbf{X}$ has compact support, $\text{supp}(\mathbb{P})$, and finite second moments. Under some assumptions on limiting behavior and smoothness on function $f$, given a $\lambda \in [0, \infty)$, define the following functional:
  \begin{align}\label{eq:pme_kappa}
\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathbb{E}\|\mathbf{X} - f(\pi_f(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2. 
  \end{align}
  The principal manifold of $\mathbf{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if
  $f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)$,
where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
\end{definition}

It is important to note that, unlike manifold learning methods that use an eigendecomposition to directly estimate coordinates in the $d$-dimensional manifold space (e.g. Isomap, locally linear embedding, and others) with only an implicit representation of the embedding function $f$, principal manifolds may be estimated via an explicit approximation of the embedding function $f$, from which the projection index $\pi_f(x)$ and the desired coordinates on the manifold can be found. In this way, regression approaches become relevant to the manifold estimation problem.

@@

\begin{figure}
  \centering
  \subfloat[\centering Data]{\label{fig:lpme_step1} \includegraphics[width=7cm]{sim_case1_plot2}}%
  \hfill
  \subfloat[\centering Step 1: Sample Size Reduction]{\label{fig:lpme_step2} \includegraphics[width=7cm]{lpme_data_reduction}}
  \vfill
  \subfloat[\centering Step 2: Initialization]{\label{fig:lpme_step3} \includegraphics[width=7cm]{lpme_initialization}}
  \hfill
  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{\label{fig:lpme_step4} \includegraphics[width=7cm]{lpme_fitting}}
  \caption{{\footnotesize LPME algorithm steps, demonstrated using longitudinal simulated data with $D=2$. Data from the functions $x_1 = r$ and $x_2 = \alpha \sin \left(\beta r + \frac{\pi}{2}\right)$ are generated at 5 time points. Random noise is added both in time and space as $\zeta \bf{g}(t) + \iota$, where $\bf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time, $\zeta$ represents a normally distributed amplitude multiplier for this translation, and $\iota$ represents normally distributed noise applied to each point individually. In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the red points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the blue surface in (d) shows the estimated smooth surface.} }
  \label{fig:lpme_steps}
\end{figure}

@@

In most practical manifold learning tasks, data are often quite large. We begin by deriving a small number of feature points in $D$-dimensional space that capture the intrinsic structure of the data cloud. This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest. Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is large. To reduce the sample size $I_t$, we use the high-dimensional mixture density estimation (HDMDE) method \citep[see][alg. 1]{mengPrincipalManifoldEstimation2021} to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Briefly, this approach is based on clustering of the data at each original time point to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using red dots in Figure \ref{fig:lpme_step2}, along with their corresponding weights, $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$, which characterize the shape of the original data. The number $N_t$ of clusters is estimated for each time point $t$ and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure. By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ \citep[see the simulation study presented in][fig. 3(c)]{mengPrincipalManifoldEstimation2021}. The centers $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$ estimated in this step are then used to reach an initial low-dimensional parameterization.

@@

Once the initial parameterization for each cluster center $\mu_{j, t}$ is obtained, the model fitting process begins by using the cluster centers $\mu_{j, t}$, weights $\hat{\theta}_{j, t}$ and the parameterizations associated with each time point to initialize a PME run. Specifically, the cluster centers $\mu_{j, t}$ and weights $\hat{\theta}_{j, t}$ replace those obtained using the HDMDE algorithm, and the Isomap-derived parameters obtained in Steps 1 and 2 of Algorithm 2 in \cite{mengPrincipalManifoldEstimation2021}, respectively.
This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa}, illustrated using multicolored lines in Figure \ref{fig:lpme_step3}. The performance of each fit $\widehat{f_t}$ is measured using $\tau_t = \frac{1}{I_t}\sum_{i=1}^{I_t}\| x_{it} - \widehat{f_t}(\pi_{\widehat{f_t}}(x_{it}))\|^{2}_{\mathbb{R}^D}$, an estimate of the mean squared distance between the observed data points and their projections to the fitted manifold $\widehat{f_t}$.

@@

A commonly used alternative approach for handling closed manifolds is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results preferably by using a technique that results in a smooth manifold in the low-dimensional space \citep[e.g.][]{mengPrincipalManifoldEstimation2021}. To avoid this process, we propose a data augmentation approach taking advantage of polar or spherical coordinates, depending on the value of $D$. This approach is motivated by the concept of ``lift" in algebraic topology \citep[sec. 1.1]{hatcher2002algebraic}. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be viewed as given in the space with $d = 1$ with the data cloud of $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.

@@

In the situations where $d = 1$, LPME is compared to the PME method naively run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point. The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package \citep{Cannoodt2018princurve}, developed using \texttt{R} \citep{rSoftware2023}. In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen. Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $N_0 = 20 \times D$, $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$. In cases where $d = 2$, LPME is again compared to the naive PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.

@@

Visualizations of the results from one example simulated case, truncated for concision, are shown in Figure \ref{fig:sim_case1}. Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold (shown in red) over time. Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.

@@

\begin{figure}
  \centering
  \includegraphics[height=9cm]{sim_case1} 
  \caption{{\footnotesize Simulation Case 1. Scatterplots of one example simulated data cloud (gray points) with $d=1$ overlaid with points on the true embedding manifold (red), the estimated LMPE function (blue), PME (green), and principal curve or surface (purple).}}
  \label{fig:sim_case1}
\end{figure}

The median and interquartile range of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Table \ref{table:simulation_results_median}. Analogous mean and standard deviation summaries of the results are given in the online supplement. Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data. Meanwhile, because the LPME approach accounts for all time points simultaneously, this approach results in lower mean squared distance values.

The result summaries indicate that in most cases, LPME provides a substantial improvement in performance over the PME and principal curve approaches when estimating the underlying manifold. As seen clearly in Figure \ref{fig:sim_case1}, while the structure of the simulated data changes noticeably between time points, the LPME estimates, shown in blue, remain relatively stable over time. This contrasts to the estimates found by the PME and principal curve approaches, which are highly sensitive to the systematic added noise in the data observed at each given time point, as expected. This ultimately results in the LPME estimates remaining closer to the true underlying manifold, shown in red. Similar results hold in most of the other simulation cases.

@@

Results of fitting the PME and LPME algorithms on the surface of the left hippocampus and left thalamus of a single participant are shown in Figure 1 in the online supplement. The previously described data augmentation approach was used to enable the fitting of closed surfaces. Visual inspection of the hippocampus estimates indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit reasonably to the data. There are two main sources of discrepancies between the data and the LPME estimate. First, at time points where the observed surface changes orientation, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points. The second difference between the LPME estimates and the observed data is seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of curves in the observed data.

Figure \ref{fig:lhipp_cross_sections} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point. These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points. This Figure reiterates that both the PME and LPME algorithms struggle to capture the distinctive sharp curves in the structure near the top and bottom corners of each plot. Additionally, we observe a gap in the estimated manifold along the lower-left edge of the surface. This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME. It also appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with high levels of curvature.

@@

While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, extension of this method to manifold estimation for groups of similar structures will be a valuable contribution to the literature allowing population-level comparisons of time-dependent trajectories of shapes. Reaching estimates associated with groups of structures raises the possibility of statistical comparisons between these groups, which, among others, could prove beneficial in a number of clinical trial settings.

>>>>

\if1\blind
{
  \title{\bf Longitudinal Principal Manifold Estimation}
  \author{
    Robert Zielinski\hspace{.2cm}\\
    Department of Biostatistics, Brown University\\
    Kun Meng \\
    Division of Applied Mathematics, Brown University\\
    and \\
    Ani Eloyan \\
    Department of Biostatistics, Brown University\\
    for the Alzheimer's Disease Neuroimaging Initiative\thanks{Data used in preparation of this article were obtained from the Alzheimer's Disease Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the investigators within the ADNI contributed to the design and implementation of ADNI and/or provided data but did not participate in analysis or writing of this report. A complete listing of ADNI investigators can be found at \href{http://adni.loni.usc.edu/wp-content/uploads/how_to_apply/ADNI_Acknowledgement_List.pdf}{\texttt{http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}}}
  \maketitle
} \fi

@@

Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}. Manifold learning approaches have been applied in various scientific fields, e.g., single-cell biology \citep{ding2023learning} and robotics \citep{gao2023k}. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures. Nonlinear manifold learning methods have been extensively studied in the literature. However, many popular approaches, such as Isomap \citep{tenenbaumGlobalGeometricFramework2000}, locally linear embedding \citep{roweisNonlinearDimensionalityReduction2000}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies. These methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension. %We use the oldest linear manifold learning method, principal component analysis \citep[PCA,][]{Pearson1901on}, as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal functional PCA method proposed by \cite{greven2011longitudinal}. As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space. However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. Additionally, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. Finally, \cite{busch2023Multiview} developed the T-PHATE manifold learning algorithm to account for the autocorrelation present in functional MRI (fMRI) data, but the temporal modeling also depends on assumptions specific to the properties of fMRI data.

@@

The framework for principal manifolds originated with the concept of principal curves \citep{hastiePrincipalCurves1989}, which are essentially curves that pass through the middle of a data cloud. Motivated by the penalization approach presented by \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework, which extends principal curves to arbitrarily high intrinsic dimensions. In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional spaces, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud ($D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest ($d=2$; for example, see Figure S1 in the online supplement). For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a $d$-variable function $u$, $\nabla^{\otimes 2} u$ denotes the Hessian matrix of $u$, defined as $r \to \left( \frac{\partial ^2 u}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for one- and higher-dimensional intrinsic spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are defined as follows.
\begin{definition}
  \label{def:principal_manifolds}
  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability distribution $\mathbb{P}$ such that $\mathbf{X}$ has compact support, $\text{supp}(\mathbb{P})$, and finite second moments. Under some assumptions on limiting behavior and smoothness on function $f$, given a $\lambda \in [0, \infty)$, define the following functional:
  \begin{align}\label{eq:pme_kappa}
\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathbb{E}\|\mathbf{X} - f(\pi_f(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2. 
  \end{align}
  The principal manifold of $\mathbf{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if
  $f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)$,
where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.
\end{definition}

It is important to note that, unlike manifold learning methods that use an eigendecomposition to directly estimate coordinates in the $d$-dimensional manifold space with only an implicit representation of the embedding function $f$ (e.g. Isomap, locally linear embedding, and others), principal manifolds may be estimated via an explicit approximation of the embedding function $f$, from which the projection index $\pi_f(x)$ and the desired coordinates on the manifold can be found. In this way, regression approaches become relevant to the manifold estimation problem.

@@

\begin{figure}
  \centering
  \subfloat[\centering Data]{\label{fig:lpme_step1} \includegraphics[width=7cm]{sim_case1_plot2}}%
  \hfill
  \subfloat[\centering Step 1: Sample Size Reduction]{\label{fig:lpme_step2} \includegraphics[width=7cm]{lpme_data_reduction}}
  \vfill
  \subfloat[\centering Step 2: Initialization]{\label{fig:lpme_step3} \includegraphics[width=7cm]{lpme_initialization}}
  \hfill
  \subfloat[\centering Steps 3 and 4: Fitting and Tuning]{\label{fig:lpme_step4} \includegraphics[width=7cm]{lpme_fitting}}
  \caption{{\footnotesize LPME algorithm steps, demonstrated using longitudinal simulated data with $D=2$. Data from the functions $x_1 = r$ and $x_2 = \alpha \sin \left(\beta r + \frac{\pi}{2}\right)$ are generated at 5 time points. Random noise is added both in time and space as $\zeta \bf{g}(t) + \iota$, where $\bf{g}(t)$ represents a normally distributed spatial translation applied to all points at each time, $\zeta$ represents a normally distributed amplitude multiplier for this translation, and $\iota$ represents normally distributed noise applied to each point individually. In (a) we show the original data cloud, in (b) we show the sample size reduction step, where the diamond-shaped points indicate the representative data to be used for the next steps of the algorithm, (c) shows the fitted time specific curves, the surface in (d) shows the estimated smooth surface.} }
  \label{fig:lpme_steps}
\end{figure}

@@

In most practical manifold learning tasks, data are often quite large. We begin by deriving a small number of feature points in $D$-dimensional space that capture the intrinsic structure of the data cloud. This approach is aimed at reducing the sample size of the original data, while maintaining the available information on the curvature of the underlying manifold of interest. Specifically, we consider $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ where the sample size $I_t$ is large. To reduce the sample size $I_t$, we use the high-dimensional mixture density estimation (HDMDE) method \citep[see][alg. 1]{mengPrincipalManifoldEstimation2021} to estimate a reduced set of points in the $D$-dimensional space that represent the shape of the data cloud $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Briefly, this approach is based on clustering of the data at each original time point to obtain cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, represented using diamond-shaped points in Figure \ref{fig:lpme_step2}, along with their corresponding weights, $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$, which characterize the shape of the original data. The number $N_t$ of clusters is estimated for each time point $t$ and is allowed to vary among time points to accommodate potential changes in the shape of the underlying structure. By implementing this approach, we reduce the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ \citep[see the simulation study presented in][fig. 3(c)]{mengPrincipalManifoldEstimation2021}. The centers $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$ estimated in this step are then used to reach an initial low-dimensional parameterization.

@@

Once the initial parameterization for each cluster center $\mu_{j, t}$ is obtained, the model fitting process begins by using the cluster centers $\mu_{j, t}$, weights $\hat{\theta}_{j, t}$ and the parameterizations associated with each time point to initialize a PME run. Specifically, the cluster centers $\mu_{j, t}$ and weights $\hat{\theta}_{j, t}$ replace those obtained using the HDMDE algorithm, and the Isomap-derived parameters obtained in Steps 1 and 2 of Algorithm 2 in \cite{mengPrincipalManifoldEstimation2021}, respectively.
This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$ corresponding to the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in \eqref{eq:newKappa}, illustrated using lines in Figure \ref{fig:lpme_step3}. The performance of each fit $\widehat{f_t}$ is measured using $\tau_t = \frac{1}{I_t}\sum_{i=1}^{I_t}\| x_{it} - \widehat{f_t}(\pi_{\widehat{f_t}}(x_{it}))\|^{2}_{\mathbb{R}^D}$, an estimate of the mean squared distance between the observed data points and their projections to the fitted manifold $\widehat{f_t}$.

@@

A commonly used alternative approach for handling closed manifolds is partitioning the high-dimensional data cloud into parts that are not self-intersecting, performing manifold fitting for each partition, and then combining the results, preferably by using a technique that results in a smooth manifold in the low-dimensional space \citep[e.g.][]{mengPrincipalManifoldEstimation2021}. To avoid this process, we propose a data augmentation approach taking advantage of polar or spherical coordinates, depending on the value of $D$. This approach is motivated by the concept of ``lift" in algebraic topology \citep[sec. 1.1]{hatcher2002algebraic}. As a simple example, the unit circle can be considered a self-intersecting manifold with $d = 1$ when viewed in two dimensions. However, by adding a third dimension equal to the angle of each observation from the origin, the manifold can be viewed as given in the space with $d = 1$ with the data cloud of $D = 3$, while avoiding any self-intersections. This enables the PME algorithm, and thus the LPME algorithm, to be fit under these circumstances. This process is illustrated graphically in Figure \ref{fig:unit_circle_augmentation}.

@@

In the situations where $d = 1$, LPME is compared to the PME method naïvely run at each time point without smoothing over time, as well as the principal curve algorithm described in \cite{hastiePrincipalCurves1989}, also run independently at each time point. The principal curve algorithm is implemented using the \texttt{principal\_curve()} function in the \texttt{princurve} package \citep{Cannoodt2018princurve}, developed using \texttt{R} \citep{rSoftware2023}. In this function, each of the three smoothing options, \texttt{smooth\_spline}, \texttt{lowess}, and \texttt{periodic\_lowess}, are tested, with the option resulting in the lowest error from the true values being chosen. Following \cite{mengPrincipalManifoldEstimation2021}, the inputs for the PME algorithm are set to $\alpha = 0.05$, and $\epsilon = 0.001$, with $\lambda_g = \exp(g)$ for $g = -15, \dots, 5$. The minimum number of cluster centers is set to $10 \times d$. In cases where $d = 2$, LPME is again compared to the naïve PME approach described above, and the principal surface estimation algorithm developed by \cite{yueParameterizationWhiteMatter2016}.

@@

Visualizations of the results from one example simulated case, truncated for concision, are shown in Figure \ref{fig:sim_case1}. Informally, the performance of each estimation method can be observed by considering the proximity of each method's estimated manifold to the true manifold over time. Thus, in Figure \ref{fig:sim_case1}, we see that the LPME-estimated manifold bears the closest resemblance to the true manifold underlying the data, while the manifolds estimated by PME and the principal curve method show greater responses to temporary random fluctuations in the data at each time point.

@@

\begin{figure}
  \centering
  \includegraphics[height=9cm]{sim_case1} 
  \caption{{\footnotesize Simulation Case 1. Scatterplots of one example simulated data cloud (circular points) with $d=1$ overlaid with points on the true embedding manifold (diamonds), the estimated LMPE function (squares), PME (triangles), and principal curve or surface (inverted triangles).}}
  \label{fig:sim_case1}
\end{figure}

The median and interquartile range of the mean squared distance from the true underlying manifold values for the estimates of each approach, as well as for the data itself, are shown in Table \ref{table:simulation_results_median}. Analogous mean and standard deviation summaries of the results are given in Table S1 of the online supplement. Because the PME, principal curve, and principal surface methods each attempt to estimate the manifold in question at each individual time point without allowing other time points to inform these estimates, they should result in similar deviations from the true underlying manifold, with differences in error resulting primarily from differing performances in fitting to the observed data. Meanwhile, because the LPME approach accounts for all time points simultaneously, this approach results in lower mean squared distance values.

The result summaries indicate that in most cases, LPME provides a substantial improvement in performance over the PME and principal curve approaches when estimating the underlying manifold. As seen clearly in Figure \ref{fig:sim_case1}, while the structure of the simulated data changes noticeably between time points, the LPME estimates remain relatively stable over time. This contrasts to the estimates found by the PME and principal curve approaches, which, as expected, are highly sensitive to the added systematic noise in the data observed at each given time point. This ultimately results in the LPME estimates remaining closer to the true underlying manifold. Similar results hold in most of the other simulation cases.

@@

Results of fitting the PME and LPME algorithms on the surface of the left hippocampus and left thalamus of a single participant are shown in Figure S1 in the online supplement. The previously described data augmentation approach was used to enable the fitting of closed surfaces. Visual inspection of the hippocampus estimates indicates that while there are slight differences between the shape shown in the data and the shape estimated by the LPME algorithm, the surface estimated by LPME appears to fit reasonably to the data. There are two main sources of discrepancies between the data and the LPME estimate. First, at time points where the observed surface changes orientation, the LPME estimates maintain a consistent orientation, reflecting the goal of encouraging stability in the structural estimates between time points. The second difference between the LPME estimates and the observed data is seen at the sharper corners of the hippocampus, where the LPME estimates do not fully capture the severity of curves in the observed data.

Figure \ref{fig:lhipp_cross_sections} depicts cross sections of the observed hippocampus surface and the estimated surface obtained using LPME and PME at each individual time point. These cross sections illustrate the differences between the estimates reached by the PME and LPME algorithms, with the LPME-based estimate being generally less responsive to changes in the shape and orientation of the hippocampus between time points. This Figure reiterates that both algorithms struggle to capture the distinctive sharp curves in the structure near the top and bottom corners of each plot. Additionally, we observe a gap in the estimated manifold along the lower-left edge of the surface. This is present for both the PME and LPME estimates, but appears to be more prevalent when using LPME. It also appears that the boundaries of the LPME-estimated surface tend to fall inside the boundaries of the PME-estimated surface, particularly at locations with high levels of curvature.

@@

While the approach described in this article is limited to the estimation of manifolds from observations taken from a single structure over time, extension of this method to manifold estimation for groups of similar structures will be a valuable contribution to the literature allowing population-level comparisons of time-dependent trajectories of shapes. Reaching estimates associated with groups of structures raises the possibility of statistical comparisons between these groups, which could prove beneficial in a number of clinical trial settings.
